---
title: "Co-expression network analysis using RNA-Seq data "
author: V. Keith Hughitt
date: "June 15, 2016"
output:
  md_document:
    toc: true
    variant: markdown_github
---

```{r knitr_settings, include=FALSE, echo=FALSE}
library('knitr')
library('knitcitations')
opts_chunk$set(fig.width=1080/96,
               fig.height=1080/96,
               fig.retina=1,
               dpi=96,
               error=FALSE)
options(digits=4)
```

Introduction
============

### Overview

Here are the main steps we are going to cover in this tutorial:

1. Data preparation
	- low-count filtering
	- log-transforming data
2. Differential expression analysis
	- filtering out non-differentially-expressed genes
3. Co-expression network construction

Things which are _not_ covered in-depth in this tutorial, but warrant
consideration: 

1. Sample quality assurance (FastQC, PCA plots, etc.)
   - [FastQC homepage](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)
   - [Bioconductor RNA-Seq workflow: Exploratory analysis and visualization](http://www.bioconductor.org/help/workflows/rnaseqGene/#exploratory-analysis-and-visualization)
2. Batch adjustment (ComBat/sva/RUVSeq)
   - [SVA tutorial](https://bioconductor.org/packages/release/bioc/vignettes/sva/inst/doc/sva.pdf)
   - [The sva package for removing batch effects and other unwanted variation in high-throughput experiments `r citep('10.1093/bioinformatics/bts034')`](http://bioinformatics.oxfordjournals.org/content/28/6/882.short)
3. Normalization (Quantile normalization/TMM/etc.)
   - [Evaluation of statistical methods for normalization and differential expression in mRNA-Seq experiments `r citep('10.1186/1471-2105-11-94')`](http://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-11-94)
   - [A comprehensive evaluation of normalization methods for Illumina high-throughput RNA sequencing data analysis `r citep('10.1093/bib/bbs046')`](http://bib.oxfordjournals.org/content/14/6/671.short)
4. A detailed discussion of the pros and cons of various approaches for
   differential expression analysis of RNA-Seq data.
   - [Comparison of software packages for detecting differential expression in RNA-seq studies `r citep('10.1093/bib/bbt086')`](http://bib.oxfordjournals.org/content/16/1/59.short)
   - [A comparison of methods for differential expression analysis of RNA-seq data `r citep('10.1186/1471-2105-14-91')`](https://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-14-91)
5. Parameter Optimization (Short solution: try out a few different ways of 
   preparing data, and measure network module enrichment each time.)
6. R programming.
   - [Coursera - R programming](https://www.coursera.org/learn/r-programming)

### Installation and Usage

### Dataset

Setup
=====

First, let's load the libraries that we will be using for this analysis. These
can all be installed using either `install.packages()`, or in some cases,
`biocLite()`.

```{r}
library('gplots')
library('ggplot2')
library('knitr')
library('limma')
library('reshape2')
library('RColorBrewer')
library('WGCNA')

# Make sure results are reproducible
set.seed(1)
```

Now, let's load the sample metadata file from the `data/` directory.

```{r load_metadata, results='asis'}
samples <- read.csv('../data/sample_metadata.csv')
kable(samples)
```

Next, we will load our RNA-Seq read counts.

```{r load_data}
raw_counts <- read.csv('../data/count_table.csv', row.names=1)
head(raw_counts)
dim(raw_counts)
```

For gene annotations, we can use the Bioconductor `Homo.sapiens` OrganismDb
package.  This meta-package combines several Human-specific annotation
packages, providing gene- and transcript-level details.

```{r}
library('Homo.sapiens')
```

OrganismDb packages can be queried in a manner similar to querying a database.
You have to specify one or more gene identifiers ('keys'), along with the type
of the identifier ('key type'), and one or more fields that you are interested
in querying.

```{r}
keytypes(Homo.sapiens)
columns(Homo.sapiens)
```

To query the package, you use the `select()` function, e.g.:

```{r}
# Example query
gene_ids <- head(keys(Homo.sapiens, keytype='ENSEMBL'), 2)
select(Homo.sapiens, keytype='ENSEMBL', keys=gene_ids, 
	   columns=c('ALIAS', 'TXCHROM', 'TXSTART', 'TXEND'))
```

Data Preparation
================

### Sample check

First, it is always wise to check the quality of your samples before continuing
with an analysis like this. There are many such checks that one can (and should)
perform, starting at the level of read quality (e.g. FastQC).

Here, we will just do a quick check using a sample-correlation heatmap.

```{r sample_heatmap}
# add a colorbar along the heatmap with sample condition
num_conditions <- nlevels(samples$condition)
pal <- colorRampPalette(brewer.pal(num_conditions, "Set1"))(num_conditions)
cond_colors <- pal[as.integer(samples$condition)]

heatmap.2(cor(raw_counts), RowSideColors=cond_colors,
		  trace='none', main='Sample correlations (raw)')
```

### Low count filtering

Now that we are satisfied with the quality of our samples, we will want to
filter our data to keep only the genes which will be informative during
differential expression analysis and network construction.

```{r}
# Remove all rows with less than n counts across all samples, where n=#samples
low_count_mask <- rowSums(raw_counts) < ncol(raw_counts)

sprintf("Removing %d low-count genes (%d remaining).", sum(low_count_mask), 
		sum(!low_count_mask))
```

### Log2 transformation

Most of the methods developed for co-expression network analysis and network 
inference were written for use with microarray data, including WGCNA!

Attempting to apply a method such as this to discrete-count RNA-Seq data
will not work out well.

There are a number of methods for working around this, in effect, making 
RNA-Seq data "look" more like microarray data, but the simplest thing is just
to _log_ the data. This will transform our discrete, over-dispersed counts
to a more Poisson-like continuous distribution.

```{r}
log_counts <- log2(raw_counts + 1)
```

Let's see how things look after logging the data.

```{r sample_density_plot}
x = melt(as.matrix(log_counts))

colnames(x) = c('gene_id', 'sample', 'value')
ggplot(x, aes(x=value, color=sample)) + geom_density()
```

```{r log2_sample_heatmap}
heatmap.2(cor(log_counts), RowSideColors=cond_colors,
		  trace='none', main='Sample correlations (log2-transformed)')
```

As you can see, after the low-count filtering and log-transformation, the 
samples within each condition are starting to behave better.

### Remove non differentially-expressed genes

Next, we will perform a series of differential expression contrasts, and use
the results to further filter out genes for which there is not a significant
amount of variance.

If you were _just_ interested in performing differential expression analysis,
this may not be the most appropriate approach. In this case, you may want to
consider additional steps such as [quantile normalization](https://en.wikipedia.org/wiki/Quantile_normalization) and/or mean-variance
adjustment with [voom](http://link.springer.com/article/10.1186/gb-2014-15-2-r29).

```{r diffexpr_filtering}
# first, let's remove any genes with _zero_ variance since these are not
# going to help us, and may cause problems with some of the models
log_counts <- log_counts[apply(log_counts, 1, var) > 0,]

# create design matrix for differential expression analysis;
# if you wanted to account for batch here, you could simply include a batch
# term in the linear model at this step, e.g.:
# mod <- model.matrix(~0+samples$condition+samples$batch)
mod <- model.matrix(~0+samples$condition)

# make model terms easier to work with
colnames(mod) <- levels(samples$condition)

fit <- lmFit(log_counts, design=mod)

# generate a list of all possible pairwise contrasts
condition_pairs <- t(combn(levels(samples$condition), 2))                                                                                                                               
                                                                                                                                                                             
comparisons <- list()                                                                                                                                          
for (i in 1:nrow(condition_pairs)) {                                                                                                                                     
    comparisons[[i]] <- as.character(condition_pairs[i,])                                                                                                      
}    

# vector to store de genes
sig_genes <- c()

# iterate over the contrasts, and perform a differential expression test for
# each pair
for (conds in comparisons) {
    # generate string contrast formula, "infLM24 - infLM4"
    contrast_formula <- paste(conds, collapse=' - ')

    contrast_mat <- makeContrasts(contrasts=contrast_formula, levels=mod)
    contrast_fit <- contrasts.fit(fit, contrast_mat)
    eb <- eBayes(contrast_fit)

    # Grab highly ranked genes; this is a pretty stringent p-value cutoff, but
    # it serves to limit the total number of genes we will use for this
    # tutorial
    sig_genes <- union(sig_genes, 
                       rownames(topTable(eb, number=Inf, p.value=0.005)))
}

# Filter out genes which were not differentially expressed for any contrast
log_counts <- log_counts[rownames(log_counts) %in% sig_genes,]
```

### Co-expression network construction

#### Construct similarity matrix

Now, we are ready to start constructing our co-expression network. The first
step is to generate a correlation, or more generally, a similarity matrix.

For this, we could use any similarity metric: Pearson correlation, Biweight
mid-correlation, mutual information, etc.

Here, we will use a metric that I have created which combines some
of the benefits from both Pearson correlation and Euclidean Distance.

```{r cor_dist}
#'
#' Similarity measure which combines elements from Pearson correlation and
#' Euclidean distance.
#' 
cordist <- function(dat) {
    cor_matrix  <- cor(t(dat))

    dist_matrix <- as.matrix(dist(dat, diag=TRUE, upper=TRUE))
    dist_matrix <- log1p(dist_matrix)
    dist_matrix <- 1 - (dist_matrix / max(dist_matrix))

    sign(cor_matrix) * ((abs(cor_matrix) + dist_matrix)/ 2)
}
```

```{r sim_matrix}
sim_matrix <- cordist(log_counts)
```

Let's see what our similarity matrix looks like at this point. Because the
`heatmap.2` function (which includes a biclustering step) can be pretty slow,
we will use a sub-sample of our data -- for visualization purposes this is
fine.

```{r sim_matrix_heatmap}
heatmap_indices <- sample(nrow(sim_matrix), 500)

heatmap.2(t(sim_matrix[heatmap_indices, heatmap_indices]),
            col=redgreen(75),
            labRow=NA, labCol=NA, 
            trace='none', dendrogram='row',
            xlab='Gene', ylab='Gene',
            main='Similarity matrix',
            density.info='none', revC=TRUE)
```

#### Construct adjacency matrix

Next, we will convert similarity matrix to an adjacency matrix.

For this, we will use the power transformation, as suggested by Zhang and
Horvath (2005).

This will help us to reduce the number of spurious correlations in the data.

Additionally, in order to separate clusters of genes which have a strong
positive correlation from those with a strong negative correlation, we will
first shift the data from the range [-1,1] to [0,1].

The combined transformation is:

$$
a_{ij} = \left(\frac{1}{2} (1 + s_{ij})\right)^\gamma
$$

Some additional reading regarding the use of power transformation:

1. [http://labs.genetics.ucla.edu/horvath/CoexpressionNetwork/Rpackages/WGCNA/faq.html](WGCNA FAQ) (E.g. item #5)
2. [https://support.bioconductor.org/p/66101/#66195](Bioconductor Support - Question: Theoretical WGCNA Question)

```{r construct_adjacency_matrix}
# Construct adjacency matrix
adj_matrix <- adjacency.fromSimilarity(sim_matrix, power=12, type='signed')

# Delete similarity matrix to free up memory
rm(sim_matrix)
gc()

# Convert to matrix
gene_ids <- rownames(adj_matrix)

adj_matrix <- matrix(adj_matrix, nrow=nrow(adj_matrix))
rownames(adj_matrix) <- gene_ids
colnames(adj_matrix) <- gene_ids
```

Same plot as before, but now for our adjacency matrix:

```{r sim_matrix_heatmap}
heatmap.2(t(adj_matrix[heatmap_indices, heatmap_indices]),
            col=redgreen(75),
            labRow=NA, labCol=NA, 
            trace='none', dendrogram='row',
            xlab='Gene', ylab='Gene',
            main='Adjacency matrix',
            density.info='none', revC=TRUE)
```

### Co-expression module detection

At this point, we already have our co-expression network. We could simply
threshold the adjacency matrix, export the remaining edges, and load it into
Cytoscape for visualization.

Before we do that, however, let's first try and detect co-expression modules in
the network.

For this, we will use hierarchical clustering following by branch-cutting.

```{r module_detection}
# Cluster gene expression profiles; the flashClust function from
# the authors of WGCNA is another options for larger datasets.
# For input, we use the reciprocal of the adjacency matrix; hierarchical
# clustering works by comparing the _distance_ between objects instead of the
# _similarity_.
gene_tree <- hclust(as.dist(1 - adj_matrix), method="average")

# we will use the cuttreeDynamicTree method to break apart the hc dendrogram
# into separate modules
module_labels <- cutreeDynamicTree(dendro=gene_tree, minModuleSize=15,
                                   deepSplit=TRUE)

# assign a color to each module for easier visualization and referencing
module_colors <- labels2colors(module_labels)
```

### Exporting the network

Another helper function for exporting our network to the metadata-rich GraphML
format...

```{r export_graphml}
#' Converts an adjaceny matrix along with some optional vertex and edge
#'  information to a GraphML graph and saves it to disk.
#'
#' @param adj_mat An n-by-n weighted or unweighted adjacency matrix normalized
#' to contain values between 0 and 1.
#' @param filename Name of file to save output to. If file already exists it
#' will be overwritten. (default: network.graphml)
#' @param weighted Whether or not the adjacency matrix should be treated as a 
#' weighted graph. (default: TRUE)
#' @param threshold For weighted networks, if a threshold value between 0 and 
#' 1 is specified, all edges with weights below that value with be dropped from
#'   the graph. (default: 0.5)
#' @param max_edge_ratio The maximum number of edges per node in the network to
#' allow. If the number of edges that would remain for the specified threshold
#' exceeds this value, the threshold will be raised to reduce the number of
#' edges remaining. (default: 3)
#' @param nodeAttr A vector with length equal to the number of vertices in the 
#' network, where the ith entry in the vector corresponds to some numeric or 
#' string annotation that should be associated with the ith node in the 
#' adjacency matrix. (default: NULL)
#' @param nodeAttrDataFrame A data frame containing one or more columns 
#' associated with the vertices in the graph.  The ith row of the dataframe 
#' should correspond to the ith entry in the adjacency matrix. (default: NULL)
#' @param edgeAttributes Extra attributes to associate with the graph edges,
#' formatted as a list of matrices of the same dimension and names as the
#' adjacency matrix.
#'
#' Examples
#' --------
#' export_network_to_graphml(adj_mat, filename='~/network.graphml',
#'                           threshold=0.3, nodeAttrDataFrame=df)
#'
#' See Also
#' --------
#' 1. http://www.inside-r.org/packages/cran/WGCNA/docs/exportNetworkToCytoscape
#' 2. http://graphml.graphdrawing.org/
#'
#' Returns
#' -------
#' An igraph graph object representing the exported graph.
export_network_to_graphml <- function (adj_mat, filename=NULL, weighted=TRUE,
                                       threshold=0.5, max_edge_ratio=3,
                                       nodeAttr=NULL, nodeAttrDataFrame=NULL,
                                       edgeAttributes=NULL, verbose=FALSE) {
    library('igraph')

    # Determine filename to use
    if (is.null(filename)) {
        filename='network.graphml'
    }

    # TODO 2015/04/09
    # Add option to rescale correlations for each module before applying
    # threshold (this is simpler than the previous approach of trying to
    # determine a different threshold for each module)
    #
    # Still, modules with very low correlations should be given somewhat
    # less priority than those with very high correlations.

    #module_colors <- unique(nodeAttrDataFrame$color)
    #module_genes <- which(nodeAttrDataFrame$color == color)
    #module_adjmat <- adj_mat[module_genes,]
    #num_genes <- length(module_genes)

    # Adjust threshold if needed to limit remaining edges
    max_edges <- max_edge_ratio * nrow(adj_mat)

    edge_to_total_ratio <- max_edges / length(adj_mat)
    edge_limit_cutoff <- as.numeric(quantile(abs(adj_mat), 1 - edge_to_total_ratio))

    # Also choose a minimum threshold to make sure that at least some edges
    # are left
    min_threshold <- as.numeric(quantile(abs(adj_mat), 0.9999))

    threshold <- min(min_threshold, max(threshold, edge_limit_cutoff))

    # Remove edges with weights lower than the cutoff
    adj_mat[abs(adj_mat) < threshold] <- 0

    # Drop any genes with no edges (TODO: Make optional?)
    orphaned <- (colSums(adj_mat) == 0)
    adj_mat <- adj_mat[!orphaned, !orphaned]

    # Also remove annotation entries
    if (!is.null(nodeAttr)) {
        nodeAttr <- nodeAttr[!orphaned]
    }
    if (!is.null(nodeAttrDataFrame)) {
        nodeAttrDataFrame <- nodeAttrDataFrame[!orphaned,]
    }

    # Keep track of non-positive edges and rescale to range 0,1
    is_zero     <- adj_mat == 0
    is_negative <- adj_mat < 0

    adj_mat <- (abs(adj_mat) - threshold) / (max(adj_mat) - threshold)
    adj_mat[is_zero] <- 0
    adj_mat[is_negative] <- -adj_mat[is_negative]

    if (verbose) {
        message(sprintf("Outputting matrix with %d nodes and %d edges", 
                        nrow(adj_mat), sum(adj_mat > 0)))
    }

    # Create a new graph and add vertices
    # Weighted graph
    if (weighted) {
        g <- graph.adjacency(adj_mat, mode='undirected', weighted=TRUE, diag=FALSE)
    } else {
        adj_mat[adj_mat != 0] <- 1
        g <- graph.adjacency(adj_mat, mode='undirected', diag=FALSE)
    }

    # Add single node annotation from vector
    if (!is.null(nodeAttr)) {
        g <- set.vertex.attribute(g, "attr", value=nodeAttr)
    }

    # Add node one or more node annotations from a data frame
    if (!is.null(nodeAttrDataFrame)) {
        for (colname in colnames(nodeAttrDataFrame)) {
            g <- set.vertex.attribute(g, colname, value=nodeAttrDataFrame[,colname])
        }
    }

    # TODO 2014/06/14
    # Add support for setting edge attributes...
    # Convert matrix of edge annotations to a list of edge pairs and values
    #x <- as.data.frame(as.table(diff))
    #E(gD)[as.character(dataSet.ext$V1) %--% as.character(dataSet.ext$V2)]$weight <- as.numeric(dataSet.ext$V3)
    #E(gD)[as.character(dataSet.ext$V1) %--% as.character(dataSet.ext$V2)]$similarity <- as.numeric(dataSet.ext$V4)
    edge_correlation_negative <- c()

    # neg_correlations[edge_list]
    edge_list <- get.edgelist(g)

    for (i in 1:nrow(edge_list)) {
        from <- edge_list[i, 1]    
        to   <- edge_list[i, 2]    
    }
    
    # Apply threshold

    # Save graph to a file
    write.graph(g, filename, format='graphml')

    # return igraph
    return(g)
}
```

```{r export_network}
# use OrganismDb to retrieve gene annotations
gene_info <- select(Homo.sapiens, keytype='ENSEMBL', keys=rownames(log_counts),
                    columns=c('TXCHROM', 'TXSTRAND', 'GENENAME'))
colnames(gene_info) <- c('gene_id', 'description', 'chr', 'strand')

# for now, just grab the description for the first transcript
gene_info <- gene_info[!duplicated(gene_info$gene_id),]

gene_info <- cbind(gene_info, module=module_colors)

# Include RGB versions of module colors for better assignment in Cytoscape
gene_info$color_rgb <- col2hex(gene_info$module)

# first, it's a good idea to check the distribution of edges weights in our
# correlation matrix. This will help us choose a reasonable cutoff for
# exporting the network.
g <- export_network_to_graphml(adj_matrix, filename='~/network.graphml',
                               threshold=0.4, nodeAttrDataFrame=gene_info)
```

References
==========

```{r results='asis', echo=FALSE}
bibliography()
```

Version Information
===================

```{r}
sessionInfo()
```
